Static and Dynamic Bethe-Salpeter Equations in the $T$-Matrix Approximation

### 2022-02-12 Manuscript/Tmatrix.tex
first complete draft

%%% FIGURE 1 %%%
\begin{figure*}
	\includegraphics[width=\textwidth]{H2}
	\caption{Singlet (left) and triplet (right) excitation energies (in \si{\eV}) of \ce{H2} as a function of the internuclear distance $R_{\ce{H-H}}$ (in \si{\angstrom}) computed at the FCI (black), BSE@$G_0W_0$ (blue), and BSE@$G_0T_0$ (red) levels with the cc-pVTZ basis.
	Raw data are reported in {\SupInf}.}
	\label{fig:H2}
\end{figure*}
%%% %%% %%% %%%

%%% FIGURE 2 %%%
\begin{figure}
	\includegraphics[width=\linewidth]{H2_gap}
	\caption{HOMO and LUMO quasiparticle energies as well as HOMO-LUMO gap (in \si{\eV}) of \ce{H2} as a function of the internuclear distance $R_{\ce{H-H}}$ (in \si{\angstrom}) computed at the $G_0W_0$ (blue) and $G_0T_0$ (red) levels with the cc-pVTZ basis.
	Raw data are reported in {\SupInf}.}
	\label{fig:H2_gap}
\end{figure}
%%% %%% %%% %%%

%%% FIGURE 3 %%%
\begin{figure*}
	\includegraphics[width=0.4\linewidth]{H2_dBSE1}
	\hspace{0.1\linewidth}
	\includegraphics[width=0.4\linewidth]{H2_dBSE3}
	\caption{
	Error with respect to FCI for the lowest singlet (left) and the lowest triplet (right) excitation energies of \ce{H2} as a function of the internuclear distance $R_{\ce{H-H}}$ (in \si{\angstrom}) computed within the static schemes (BSE@$G_0W_0$ and BSE@$G_0T_0$) and the dynamically-corrected schemes (dBSE@$G_0W_0$ and dBSE@$G_0T_0$) .
	The cc-pVTZ basis is employed for all calculations.
	Raw data are reported in {\SupInf}.}
	\label{fig:H2_dBSE}
\end{figure*}
%%% %%% %%% %%%

@@

%%% FIGURE 4 %%%
\begin{figure*}
	\includegraphics[width=\textwidth]{BeH2}
	\caption{Singlet (left) and triplet (right) excitation energies (in \si{\eV}) of \ce{BeH2} as a function of the distance $R_{\ce{Be-H}}$ (in \si{\angstrom}) computed at the FCI (black), BSE@$G_0W_0$ (blue), and BSE@$G_0T_0$ (red) levels with the cc-pVDZ basis.
	The dynamically-corrected BSE excitation energies are represented as thin lines for dBSE@$G_0W_0$ (blue) and dBSE@$G_0T_0$ (red).
	Raw data are reported in {\SupInf}.}
	\label{fig:BeH2}
\end{figure*}
%%% %%% %%% %%%

%%% FIGURE 5 %%%
\begin{figure}
	\includegraphics[width=\linewidth]{BeH2_gap}
	\caption{HOMO and LUMO quasiparticle energies as well as HOMO-LUMO gap (in \si{\eV}) of \ce{BeH2} as a function of the internuclear distance $R_{\ce{Be-H}}$ (in \si{\angstrom}) computed at the BSE@$G_0W_0$ (blue) and BSE@$G_0T_0$ (red) levels with the cc-pVDZ basis.
	Raw data are reported in {\SupInf}.}
	\label{fig:BeH2_gap}
\end{figure}
%%% %%% %%% %%%

@@

As a third and final example, we compute the excitation energies associated with the two lowest singlet and two lowest triplet excited states of water at equilibrium geometry (see Fig.~\ref{fig:H2O}).
Note that all these excited states are of Rydberg nature and correspond to $n \to 3s$ and $n \to 3p$ transitions for the $B_1$ and $A_2$ states, respectively.
In addition to the BSE-based models studied in the present manuscript, we have selected well-known wave function methods, namely CIS, CIS(D), TDHF, and FCI (taken as reference) and computed the excitation energies for these transitions.
Two key observations can be made:
i) BSE@$G_0W_0$ is by far the best performer with a slight overestimation of the order of \SI{0.1}{eV} (as compared to FCI);
ii) BSE@$G_0T_0$ systematically underestimates the excitation energies [similarly to CIS(D)] and outperforms CIS, CIS(D) and TDHF for the singlet states only. 
These general trends are also observed for other systems and they nicely evidence the crucial role of the screening in $GW$, hinting that a screened version of the $T$-matrix formalism as proposed in Ref.~\onlinecite{Romaniello_2012} might be a promising way for improvement.

@@

%%%%%%%%%%%%%%%%%%%%%%
\section{Conclusion}
\label{sec:ccl}
%%%%%%%%%%%%%%%%%%%%%%
To conclude, the BSE@$GT$ formalism seems to be, overall, less accurate than BSE@$GW$ in the context of the computation of molecular excitation energies.
However, it still outperforms conventional methods such CIS and TDHF for singlet states.
It would be interesting to investigate its performance for the computation of ground-state correlation energies within the adiabatic connection fluctuation dissipation formalism where BSE@$GW$ has been shown to be particularly outstanding. \cite{Maggio_2016,Holzer_2018b,Loos_2020e}
The combination of $GT$ and $GW$ via the range separation of the Coulomb operator to avoid double counting of the low-order diagrams is also a promising avenue.
Work along these lines are currently under progress.
Finally, the unrestricted and spin-flip extensions of the present formalism are currently being developed.

@@

In order to derive the dynamical kernel $\tilde{T}_{ib,aj}$ given in equation Eq.~\eqref{eq:dT}, we follow Ref.~\onlinecite{Strinati_1980} (see also Ref.~\onlinecite{Loos_2020e}) and start from the equation for the BSE right-hand side electron-hole amplitude 

@@

\begin{align}

@@

	\Xi(3,5,4,6) & =  i\frac{\delta \Sigma(3,4)}{\delta G(6,5)}\approx -T(3,5,4,6).
\end{align}
Equation (\ref{eq:Strinati_1}) is derived by assuming a particular (resonant) pole $\omega_S=E_S-E_0>0$ of $L$ to be isolated from the other poles, which is usually the case for neutral excitations in finite systems, and  by assuming the poles of $L_0$ to be different form $\omega_S$, which is in general the case.

@@

where $v$ is the bare Coulomb operator and we neglect the functional derivative $\delta T/\delta G$ in the kernel $\Xi$. We also notice that the first two terms on the right-hand side of Eq.~(\ref{eq:T-matrix}) Hartree and exchange contributions to the T matrix ($T^{\text{HF}}$), whereas the last term is the correlation contribution $T^c=T-T^{\text{HF}}$.

@@

	\Theta(\pm\tau)e^{-i\alpha \tau}=\mp\frac{1}{2\pi i}\lim_{\eta\to 0^+}\int d\omega \frac{1}{\omega-\alpha\pm i\eta}e^{-i\omega\tau}.

@@

For the resonant case, \ie, $\omega_S>0$, 
\begin{equation}
\label{eq:chi_tau}
	\chi_S(\bx_1,\bx_{1'},\tau_1) 
	=-e^{i\omega_S\abs{\tau_1}/2} \sum_{jb}\SO{b}(\bx_1)\SO{j}(\bx_{1'}) \mel{N}{\hat{c}^\dagger_{j}\hat{c}_b}{N,S}
  	\qty[\Theta(\tau_1)e^{-i\e{b}{}\tau_1}+\Theta(-\tau_1)e^{-i\e{j}{}\tau_1} ]
\end{equation}
  where $\hat{c}^\dagger_{p}$ and $\hat{c}_p$ are the usual creation and annihilation operators, respectively, and $\ket{N}$ and $\ket{N,S}$ are the ground state and the $S$th excited state, respectively, of the $N$-electron system.
After some algebraic steps, one gets
\begin{multline}
	-\qty(\e{a}{}-\e{i}{}-\omega_S)\mel{N}{\hat{c}^\dagger_{i}\hat{c}_a}{N,S}
	\\
    = -\sum_{jb}\mel{N}{\hat{c}^\dagger_{j}\hat{c}_b}{N,S}
    \qty{ -\frac{i}{2\pi}\int d\omega \lim_{\eta \to 0^+}\ \Bar{T}_{ib,aj}(\omega) e^{-i\omega\eta} \qty[\frac{1}{\omega_S-\omega+\e{j}{}+\e{i}{}+i\eta}+\frac{1}{\omega_S+\omega-\e{b}{}-\e{a}{}+i\eta} ] }
\end{multline}
where we have defined 
\begin{equation}
	\Bar{T}_{ib,aj}(\tau_{34})=\int d\bx_3 d\bx_4 d\bx_5 d\bx_6 \SO{a}(\bx_3) \SO{i}(\bx_4)\Bar{T}(\bx_3,\bx_5,\bx_4,\bx_6;\tau_{34}) \SO{b}(\bx_6)\SO{j}(\bx_5).
\end{equation}
Using the definition $X_{qp,S}=\mel{N}{\hat{c}^\dagger_{q}\hat{c}_p}{N,S}$, we arrive at

\begin{equation}
	\label{eq:egval}
	\qty(\e{a}{}-\e{i}{}-\omega_S) X_{ia,S}+\sum_{jb}X_{jb,S}\mel{ib}{}{aj}+\sum_{jb}X_{jb,S}\widetilde{T}_{ib,aj}(\omega_S)=0
\end{equation}
%with $$\langle ab|ij\rangle=\int d\bx_1d\bx_2 \SO{a}(\bx_1) \SO{b}(\bx_2)v(\bx_1,\bx_2) \SO{i}(\bx_1) \SO{j}(\bx_2),$$ 
where the spectral representation of the dynamical $T$-matrix is
\begin{equation}
	\widetilde{T}_{ib,aj}(\omega_S)
	= \frac{i}{2\pi}\int d\omega \lim_{\eta \to 0^+}\Bar{T}^c_{ib,aj}(\omega)e^{-i\omega\eta} \left[\frac{1}{\omega_S-\omega+\e{j}{}+\e{i}{}+i\eta}+\frac{1}{\omega_S+\omega-\e{b}{}-\e{a}{}+i\eta}\right]. 
 \label{eq:H2preso_eigen_2_T}
\end{equation}
Equation \eqref{eq:egval} represents a non-linear eigenvalue equation to calculate the positive excitation energies of a system, which can be rewritten as
\begin{equation}
\label{eq:H2preso_eigen}
	A_{ia,jb}(\omega_S) X_{jb,S}=\omega_S X_{ia,S}
\end{equation}
with
\begin{equation}
	A_{ia,jb}(\omega_S)=(\e{a}{}-\e{i}{})\delta_{ij}\delta_{ab} +\mel{ib}{}{aj}+\widetilde{T}_{ib,aj}(\omega_S).
\end{equation}
%Here $H_{ia,jb}(\omega_S)=A_{ia,jb}(\omega_S)$ and $H_{ia,bj}(\omega_S)=B_{ia,jb}(\omega_S)$.
If one drops the dynamical part $\widetilde{T}$, one ends up with the usual time-dependent Hartree-Fock (TDHF) equations. 
To calculate the correlation contribution, one must employ \eqref{eq:T} in Eq.~\eqref{eq:H2preso_eigen_2_T}, and after integration over the frequency, it yields Eq.~\eqref{eq:dT}.

>>>>

%%% FIGURE 1 %%%
\begin{figure*}[t]
	\includegraphics[width=\textwidth]{H2}
	\caption{Singlet (left) and triplet (right) excitation energies (in \si{\eV}) of \ce{H2} as a function of the internuclear distance $R_{\ce{H-H}}$ (in \si{\angstrom}) computed at the FCI (black), BSE@$G_0W_0$ (blue), and BSE@$G_0T_0$ (red) levels with the cc-pVTZ basis.
	Raw data are reported in {\SupInf}.}
	\label{fig:H2}
\end{figure*}
%%% %%% %%% %%%

%%% FIGURE 2 %%%
\begin{figure}[t]
	\includegraphics[width=\linewidth]{H2_gap}
	\caption{HOMO and LUMO quasiparticle energies as well as HOMO-LUMO gap (in \si{\eV}) of \ce{H2} as a function of the internuclear distance $R_{\ce{H-H}}$ (in \si{\angstrom}) computed at the $G_0W_0$ (blue) and $G_0T_0$ (red) levels with the cc-pVTZ basis.
	Raw data are reported in {\SupInf}.}
	\label{fig:H2_gap}
\end{figure}
%%% %%% %%% %%%

%%% FIGURE 3 %%%
\begin{figure*}[t]
	\includegraphics[width=0.4\linewidth]{H2_dBSE1}
	\hspace{0.1\linewidth}
	\includegraphics[width=0.4\linewidth]{H2_dBSE3}
	\caption{
	Error with respect to FCI for the lowest singlet (left) and the lowest triplet (right) excitation energies of \ce{H2} as a function of the internuclear distance $R_{\ce{H-H}}$ (in \si{\angstrom}) computed within the static schemes (BSE@$G_0W_0$ and BSE@$G_0T_0$) and the dynamically-corrected schemes (dBSE@$G_0W_0$ and dBSE@$G_0T_0$) .
	The cc-pVTZ basis is employed for all calculations.
	Raw data are reported in {\SupInf}.}
	\label{fig:H2_dBSE}
\end{figure*}
%%% %%% %%% %%%

@@

%%% FIGURE 4 %%%
\begin{figure*}[t]
	\includegraphics[width=\textwidth]{BeH2}
	\caption{Singlet (left) and triplet (right) excitation energies (in \si{\eV}) of \ce{BeH2} as a function of the distance $R_{\ce{Be-H}}$ (in \si{\angstrom}) computed at the FCI (black), BSE@$G_0W_0$ (blue), and BSE@$G_0T_0$ (red) levels with the cc-pVDZ basis.
	The dynamically-corrected BSE excitation energies are represented as thin lines for dBSE@$G_0W_0$ (blue) and dBSE@$G_0T_0$ (red).
	Raw data are reported in {\SupInf}.}
	\label{fig:BeH2}
\end{figure*}
%%% %%% %%% %%%

%%% FIGURE 5 %%%
\begin{figure}[t]
	\includegraphics[width=\linewidth]{BeH2_gap}
	\caption{HOMO and LUMO quasiparticle energies as well as HOMO-LUMO gap (in \si{\eV}) of \ce{BeH2} as a function of the internuclear distance $R_{\ce{Be-H}}$ (in \si{\angstrom}) computed at the BSE@$G_0W_0$ (blue) and BSE@$G_0T_0$ (red) levels with the cc-pVDZ basis.
	Raw data are reported in {\SupInf}.}
	\label{fig:BeH2_gap}
\end{figure}
%%% %%% %%% %%%

@@

As a third and final example, we compute the excitation energies associated with the two lowest singlet and two lowest triplet excited states of water at equilibrium geometry (see Fig.~\ref{fig:H2O}).
Note that all these excited states are of Rydberg nature and correspond to $n \to 3s$ and $n \to 3p$ transitions for the $B_1$ and $A_2$ states, respectively. \cite{Loos_2018a}
In addition to the BSE-based models studied in the present manuscript, we have selected well-known wave function methods, \cite{Bene_1971,Head-Gordon_1994,Head-Gordon_1995,Dreuw_2005} namely CIS, CIS(D), TDHF, and FCI (taken as reference) and computed the excitation energies for these transitions.
Two key observations can be made:
i) BSE@$G_0W_0$ is by far the best performer with a slight overestimation of the order of \SI{0.1}{eV} (as compared to FCI);
ii) BSE@$G_0T_0$ systematically underestimates the excitation energies [similarly to CIS(D)] and outperforms CIS, CIS(D) and TDHF for the singlet states only. 
These general trends are also observed for other systems and they nicely evidence the crucial role of the screening in $GW$, hinting that a screened version of the $T$-matrix formalism as proposed in Ref.~\onlinecite{Romaniello_2012} might be a promising way for improvement.

@@

%%%%%%%%%%%%%%%%%%%%%%
\section{Conclusion}
\label{sec:ccl}
%%%%%%%%%%%%%%%%%%%%%%
We have derived and implemented, for the first time, the static and dynamic Bethe-Salpeter equations when one considers $T$-matrix quasiparticle energies as well as a $T$-matrix-based kernel.
The performance of the static scheme and its perturbative dynamical correction have been assessed by computing the neutral excited states of several molecular systems.
Our results suggest that the BSE@$GT$ formalism seems to be, overall, less accurate than BSE@$GW$ in the context of the computation of molecular excitation energies.
However, it still outperforms conventional methods such CIS and TDHF for singlet states.
It would be interesting to investigate its performance for the computation of ground-state correlation energies within the adiabatic connection fluctuation dissipation formalism where BSE@$GW$ has been shown to be particularly outstanding. \cite{Maggio_2016,Holzer_2018b,Loos_2020e}
The combination of $GT$ and $GW$ via the range separation of the Coulomb operator to avoid double counting of the low-order diagrams is also a promising avenue.
Work along these lines are currently under progress.
Finally, the unrestricted and spin-flip extensions of the present formalism are currently being developed.

@@

In order to derive the dynamical kernel $\tilde{T}_{ib,aj}$ given in Eq.~\eqref{eq:dT}, we follow Ref.~\onlinecite{Strinati_1980} (see also Ref.~\onlinecite{Loos_2020e}) and start \titou{from the equation for the BSE right-hand side electron-hole amplitude?!}

@@

@@

	\Xi(3,5,4,6) =  i\frac{\delta \Sigma(3,4)}{\delta G(6,5)}\approx -T(3,5,4,6)
Equation (\ref{eq:Strinati_1}) is derived by assuming that i) the (resonant) pole $\omega_S=E_S-E_0>0$ of $L$ is isolated from the other poles (which is usually the case for neutral excitations in finite systems), and  ii) the poles of $L_0$ are different form $\omega_S$ (which is also generally the case).

@@

where $v$ is the bare Coulomb operator and we neglect the functional derivative $\delta T/\delta G$ in the kernel $\Xi$. 
The first two terms in the right-hand side of Eq.~\eqref{eq:T-matrix} are the Hartree and exchange contributions to the $T$-matrix, whereas the last term is the correlation contribution.

@@

	\Theta(\pm\tau)e^{-i\alpha \tau}=\mp\frac{1}{2\pi i}\lim_{\eta\to 0^+}\int d\omega \frac{1}{\omega-\alpha\pm i\eta}e^{-i\omega\tau}

@@

For the resonant case, \ie, $\omega_S>0$, 
\begin{equation}
\label{eq:chi_tau}
	\chi_S(\bx_1,\bx_{1'},\tau_1) 
	=-e^{i\omega_S\abs{\tau_1}/2} \sum_{jb}\SO{b}(\bx_1)\SO{j}(\bx_{1'}) \mel{N}{\hat{c}^\dagger_{j}\hat{c}_b}{N,S}
  	\qty[\Theta(\tau_1)e^{-i\e{b}{}\tau_1}+\Theta(-\tau_1)e^{-i\e{j}{}\tau_1} ]
\end{equation}
  where $\hat{c}^\dagger_{p}$ and $\hat{c}_p$ are the usual creation and annihilation operators, respectively, and $\ket{N}$ and $\ket{N,S}$ are the ground state and the $S$th excited state, respectively, of the $N$-electron system.
After some algebraic steps, one gets
\begin{multline}
	-\qty(\e{a}{}-\e{i}{}-\omega_S)\mel{N}{\hat{c}^\dagger_{i}\hat{c}_a}{N,S}
	\\
    = -\sum_{jb}\mel{N}{\hat{c}^\dagger_{j}\hat{c}_b}{N,S}
    \qty{ -\frac{i}{2\pi}\int d\omega \lim_{\eta \to 0^+}\ \Bar{T}_{ib,aj}(\omega) e^{-i\omega\eta} \qty[\frac{1}{\omega_S-\omega+\e{j}{}+\e{i}{}+i\eta}+\frac{1}{\omega_S+\omega-\e{b}{}-\e{a}{}+i\eta} ] }
\end{multline}
where we have defined 
\begin{equation}
	\Bar{T}_{ib,aj}(\tau_{34})=\int d\bx_3 d\bx_4 d\bx_5 d\bx_6 \SO{a}(\bx_3) \SO{i}(\bx_4)\Bar{T}(\bx_3,\bx_5,\bx_4,\bx_6;\tau_{34}) \SO{b}(\bx_6)\SO{j}(\bx_5).
\end{equation}
Using the definition $X_{ia,S}=\mel{N}{\hat{c}^\dagger_{i}\hat{c}_a}{N,S}$, we arrive at

\begin{equation}
	\label{eq:egval}
	\qty(\e{a}{}-\e{i}{}-\omega_S) X_{ia,S}+\sum_{jb}X_{jb,S}\mel{ib}{}{aj} \titou{-} \sum_{jb}X_{jb,S}\widetilde{T}_{ib,aj}(\omega_S)=0
\end{equation}
%with $$\langle ab|ij\rangle=\int d\bx_1d\bx_2 \SO{a}(\bx_1) \SO{b}(\bx_2)v(\bx_1,\bx_2) \SO{i}(\bx_1) \SO{j}(\bx_2),$$ 
where the spectral representation of the dynamical $T$-matrix is
\begin{equation}
	\widetilde{T}_{ib,aj}(\omega_S)
	= \frac{i}{2\pi}\int d\omega \lim_{\eta \to 0^+}\Bar{T}^c_{ib,aj}(\omega)e^{-i\omega\eta} \left[\frac{1}{\omega_S-\omega+\e{j}{}+\e{i}{}+i\eta}+\frac{1}{\omega_S+\omega-\e{b}{}-\e{a}{}+i\eta}\right]. 
 \label{eq:H2preso_eigen_2_T}
\end{equation}
Equation \eqref{eq:egval} represents a non-linear eigenvalue equation to calculate the positive excitation energies of a system, which can be rewritten as
\begin{equation}
\label{eq:H2preso_eigen}
	\titou{\sum_{jb}} A_{ia,jb}(\omega_S) X_{jb,S}=\omega_S X_{ia,S}
\end{equation}
with
\begin{equation}
	A_{ia,jb}(\omega_S)=(\e{a}{}-\e{i}{})\delta_{ij}\delta_{ab} + \mel{ib}{}{aj} \titou{-} \widetilde{T}_{ib,aj}(\omega_S).
\end{equation}
%Here $H_{ia,jb}(\omega_S)=A_{ia,jb}(\omega_S)$ and $H_{ia,bj}(\omega_S)=B_{ia,jb}(\omega_S)$.
If one drops the dynamical part $\widetilde{T}$, one ends up with the usual time-dependent Hartree-Fock (TDHF) equations. \cite{Dreuw_2005}
To calculate the correlation contribution, one must employ Eq.~\eqref{eq:T} in Eq.~\eqref{eq:H2preso_eigen_2_T}, and after integration over the frequency, it yields Eq.~\eqref{eq:dT}.
